---
editor: 
  markdown: 
    wrap: 72
---

# Lab in R {#sec-pointslabR .unnumbered}

## Points

Think about the last time you walked through a city. Some things around you — lamp posts, bus stops, benches — sit exactly where they were put, and stay there. Others — where a crime happened, where someone hailed a taxi, where a photo was taken — could have happened *almost anywhere*, but happened to happen *right there*. Points, it turns out, can be read in two completely different ways.

::: {.callout-note}
## Two ways to think about a point

**Fixed objects** — the location is just a given fact (a bus stop is where it is). Analysing these is a lot like analysing polygons or lines: the geometry describes something that already exists.

**Events** — the location is the *interesting* part. The thing could theoretically have happened anywhere, but it manifested here rather than there. This is the lens we'll use for the rest of this notebook.
:::

When we zoom out from a single event to a whole collection of them, we get a **point pattern** — and the *arrangement* of those points becomes the object of study in its own right.

> 🔎 Think of crime in a city. In principle, a crime could happen on almost any street corner. In practice, they cluster — some blocks see them constantly, others almost never. That clustering (or lack of it) is exactly what point pattern analysis is built to describe.

Point patterns come in two flavours:

- **Unmarked** — all you have is *where*. Just the coordinates of each crime, nothing else.
- **Marked** — you also know *what*. The type of crime, the damage caused, the time of day — extra attributes riding along with each location.

::: {.callout-tip}
## The questions driving this notebook

- What's the *shape* of the distribution — clustered, dispersed, random?
- Is there structure we can actually detect statistically, or does it just look like a pattern to the human eye?
- *Why* here and not there? What process could be generating what we see?
:::

This notebook is a gentle, hands-on introduction to working with point patterns in `R` — reading them in, transforming them, and building up a toolkit of ways to visualize what they're telling you.

## Installing Packages

```{r setup, message=FALSE, warning=FALSE}
# Check if installed and load
# Simple Features, used for working with spatial data.
library(sf)         # Simple Features, spatial data manipulation
library(dplyr)      # Data manipulation and transformation
library(ggplot2)    # Data visualization
library(basemapR)   # Static maps
library(viridis)    # Color palettes for data visualization
library(factoextra) # Visualising clustering results

# Clustering and Dimensionality Reduction
library(dbscan)     # Density-based spatial clustering
```

## Data

We are going to continue with Airbnb data in a different part of the
world.

### Airbnb Buenos Aires

Let's read in the point dataset:

```{r}
# read the AirBnb listing
listings <- read.csv("data/BuenosAires/listings_nooutliers.csv")

summary(listings)
```

Let's finish preparing it:

```{r}
# locate the longitude and latitude
names(listings)

listings <- listings %>%
  st_as_sf(coords = c(8, 7)) %>% # create pts from coordinates
  st_set_crs(4326) # set the crs
```

### Adminstrative Areas

We will later use administrative areas for aggregation. Let's load them.

```{r}
BA <- read_sf("data/BuenosAires/neighbourhoods_BA.shp") #read shp

BA <- st_make_valid(BA) # make geometry valid
```

### Spatial Join

```{r spatialjoin}
# spatial overlay between points and polygons
listings_BA <- st_join(BA, listings)

# read the first lines of the attribute table
head(listings_BA)
```

## One-to-one

The first approach we review here is the one-to-one approach, where we
place a dot on the screen for every point to visualise. We are going to
use `ggplot` and plot the points by neighbourhood.

```{r}
ggplot() +
  geom_sf(data = listings, aes(color = neighbourhood))
```

We can visualise a bit better with a basemap

::: callout-important
CARTO retired free, unauthenticated access to their tiles (`dark`,
`positron`, `voyager`, `hydda`) — using them without your own CARTO API
key now returns a tile stamped "API KEY REQUIRED". `google-*` styles
need their own separate Google Maps API key too. `mapnik` (plain
OpenStreetMap tiles) stays free and key-free.
:::

```{r}
ggplot() +
  base_map(st_bbox(listings), basemap = 'mapnik', increase_zoom = 2) + 
  geom_sf(data = listings, aes(color = neighbourhood), size = 0.5, alpha = 0.5) +
  theme(legend.position = "none")

```

## Points meet polygons

The approach presented above works until a certain number of points to
plot; tweaking dot transparency and size only gets us so far and, at
some point, we need to shift the focus. Having learned about visualizing
lattice (polygon) data, an option is to "turn" points into polygons and
apply techniques like choropleth mapping to visualize their spatial
distribution. To do that, we will overlay a polygon layer on top of the
point pattern, join the points to the polygons by assigning to each
point the polygon where they fall into, and create a choropleth of the
counts by polygon.

This approach is intuitive but of course raises the following question:
what polygons do we use to aggregate the points? Ideally, we want a
boundary delineation that matches as closely as possible the point
generating process and partitions the space into areas with a similar
internal intensity of points. However, that is usually not the case, no
less because one of the main reasons we typically want to visualize the
point pattern is to learn about such generating process, so we would
typically not know a priori whether a set of polygons match it. If we
cannot count on the ideal set of polygons to begin with, we can adopt
two more realistic approaches: using a set of pre-existing irregular
areas or create a artificial set of regular polygons. Let's explore
both.

### Irregular lattices

To exemplify this approach, we will use the administrative areas we have
loaded above. Let's add them to the figure above to get better context
(unfold the code if you are interested in seeing exactly how we do
this):

```{r}
ggplot() +
  geom_sf(data = BA, fill = NA, size = 1.5) +
  geom_sf(data = listings, aes(color = neighbourhood), size = 0.5, alpha = 0.5) +
  theme(legend.position="none")
```

Now we need to know how many airbnb each area contains. Our airbnb table
already contains the neighbourhood ID following our use of `st_join`.
Now, all we need to do is counting by area and attaching the count to
the areas table. We can also calculate the mean price of each area.

We rely here on the `group_by` function which takes all the airbnbs in
the table and *group* them *by* neighbourhood. Once grouped, we apply
function `n()`, which counts how many elements each group has and
returns a column indexed on the neighbourhood level with all the counts
as its values. We end by assigning the counts to a newly created column
in the table.

This chunk may take a bit longer to run:

```{r aggregate}
# aggregate at district level
airbnb_neigh_agg <- listings_BA %>% 
  group_by(neighbourh) %>% # group at neighbourhood level
  summarise(count_airbnb = n(),  # create count
            mean_price = mean(price)) # average price

head(airbnb_neigh_agg)
```

The lines above have created a new column in our table called
`count_airbnb` that contains the number of airbnb that have been taken
within each of the polygons in the table. `mean_price` shows the mean
price per neighbourhood.

At this point, we are ready to map the counts. Technically speaking,
this is a choropleth just as we have seen many times before:

```{r mapoutput2}
map_BA <- ggplot()+
  geom_sf(data = airbnb_neigh_agg, inherit.aes = FALSE, aes(fill = count_airbnb), colour = "white") + 
  scale_fill_viridis("Count", direction = -1, option = "viridis" ) + 
  ggtitle("Count of Airbnbs by Neighbourhood") +
  geom_sf_text(data = airbnb_neigh_agg,
               aes(label = neighbourh),
               fun.geometry = sf::st_centroid, size=2) +
  theme_void()

map_BA
```

The map above clearly shows a concentration of airbnb in the
neighbourhoods of Palermo and Recoleta. However, it is important to
remember that the map is showing raw counts. In the case of airbnbs, as
with many other phenomena, it is crucial to keep in mind the "container
geography" (MAUP). In this case, different administrative areas have
different sizes. Everything else equal, a larger polygon may contain
more photos, simply because it covers a larger space. To obtain a more
accurate picture of the intensity of photos by area, what we would like
to see is a map of the density of photos, not of raw counts. To do this,
we can divide the count per polygon by the area of the polygon.

Let's first calculate the area in Sq. metres of each administrative
delineation:

```{r}
# Calculate area in square kilometers and add it to the data frame
airbnb_neigh_agg <- airbnb_neigh_agg %>%
  mutate(area_km2 = as.numeric(st_area(.) / 1e6)) # 1e6 just means 1000000

# Calculate density
airbnb_neigh_agg <- airbnb_neigh_agg %>%
  mutate(density = count_airbnb / area_km2)

```

With the density at hand, creating the new choropleth is similar as
above:

```{r, collapse=TRUE}
map_BA_density <- ggplot()+
  geom_sf(data = airbnb_neigh_agg, inherit.aes = FALSE, aes(fill = density), colour = "white") +
  scale_fill_viridis("Density", direction = -1, option = "viridis") + 
  ggtitle("Density of Airbnbs by Neighbourhood") +
  geom_sf_text(data = airbnb_neigh_agg,
               aes(label = neighbourh),
               fun.geometry = sf::st_centroid, size=2) +
  theme_void()

map_BA_density

```

We can see some significant differences. Why is that? Have a chat with
the person next to you.

### Regular lattices: hex-binning

Sometimes we either do not have any polygon layer to use or the ones we
have are not particularly well suited to aggregate points into them. In
these cases, a sensible alternative is to create an artificial topology
of polygons that we can use to aggregate points. There are several ways
to do this but the most common one is to create a grid of hexagons. This
provides a regular topology (every polygon is of the same size and
shape) that, unlike circles, cleanly exhausts all the space without
overlaps and has more edges than squares, which alleviates edge
problems.

::: callout-important
If you are still still not sure on the difference between geographic
coordinate systems and projected coordinated system go back to Lecture
1.
:::

First we need to make sure we are in a projected coordinated system:

```{r}
BA_proj <- st_transform(BA, 22176) # CRS for Argentina

listings_proj <- st_transform(listings, st_crs(BA_proj)) # Making sure both files have the same crs

# You can plot to check the data overlaps correctly
# plot(BA$geometry)
# plot(listings$geometry, add=TRUE)
```

Then we create a grid that's 500 metre by 500 metres. We need to be in a
projected coordinate system for this to work.

```{r}
grid <- st_make_grid(
   BA_proj,
   cellsize = 500,
   crs = 4326, 
   what = "polygons",
   square = FALSE) # creation of grid

plot(grid) # plot grid
```

To avoid any issues, convert the grid to an `sf` objects, which extends
`data.frame`-like objects with a simple feature list column.

```{r}
# Convert 'grid' to a simple features object
grid <- st_sf(grid)

# Add a new column 'n_airbnb' with counts of intersections with 'listings_proj'
grid <- grid %>% mutate(n_airbnb = lengths(st_intersects(grid, listings_proj)))

# Filter rows where 'n_airbnb' is greater than 1
grid_filtered <- filter(grid, n_airbnb > 1)
```

Let's unpack the coded here:

- `mutate` : This is a function from the `dplyr` package we've used many
  times before and it is used to add new variables or modify existing
  variables in a data frame.
- `lengths`: This function is used to compute the lengths of the
  elements in a list. Here, it is applied to the result of the
  `st_intersects` function.
- `st_intersects`(grid, listings_proj): This is a spatial operation
  using the `sf` package. It checks for spatial intersections between
  the objects in the grid and listings_proj data frames. This operation
  returns a list where each element corresponds to the intersections of
  a grid cell with the listings.
- The result of `st_intersects` is passed to `lengths` to determine the
  number of intersections for each grid cell.

The final result is that a new variable n_airbnb is added to the grid
data frame, representing the number of intersections (or occurrences)
for each grid cell with the listings in listings_proj.

```{r message=FALSE}
ggplot() +
  geom_sf(data = grid_filtered, aes(fill = n_airbnb), color = "white", linewidth = 0.1) +
  scale_fill_viridis("Airbnb\ncounts", direction = -1, trans = "log10") +
  ggtitle("Hex-binned Airbnb counts") +
  theme_void()
```

## Kernel Density Estimation

Hex-binning is a quick fix when you don't have a sensible polygon layer
to aggregate into. But it doesn't escape the
[modifiable areal unit problem](mapvector.html) — we're still drawing
arbitrary boundaries and counting inside them, so the result can still
mismatch the underlying pattern.

**Kernel density estimation (KDE)** avoids the problem entirely by never
aggregating into areas at all. Instead of asking *"how many points fell
inside this box?"*, KDE asks *"how much point-ness is there at this
exact spot?"* — counting nearby points more heavily than distant ones,
and producing a smooth continuous surface rather than a set of bins.

::: {.callout-note}
## The one parameter that matters: bandwidth

Bandwidth controls how far each point's influence spreads.

-   **Small bandwidth** → a spiky surface that tracks individual points.
    Lots of detail, lots of noise.
-   **Large bandwidth** → a smooth blob. Clean, but real local structure
    gets washed out.

There is no single "correct" value. Choosing one is a judgement call
about what scale of pattern you're trying to show.
:::

### Kernel densities with `ggplot`

The good news: you don't need any extra packages for this. `ggplot2` can
compute a KDE directly with `geom_density_2d_filled()`, which is the
same approach the Python version of this lab uses via `seaborn`.

It needs plain x/y columns rather than an `sf` geometry column, so we
pull the coordinates out first. Note we use `listings_proj` (projected,
in metres) rather than the unprojected version — distances need to be
meaningful for the smoothing to make sense.

```{r}
listings_xy <- listings_proj |>
  st_coordinates() |>
  as.data.frame()

head(listings_xy)
```

Now we can map it:

```{r}
ggplot(listings_xy, aes(x = X, y = Y)) +
  geom_density_2d_filled(bins = 12) +
  scale_fill_viridis_d(name = "Density") +
  coord_sf() +
  ggtitle("KDE of Airbnbs in Buenos Aires") +
  theme_void()
```

Let's unpack that:

-   `geom_density_2d_filled()` does the density estimation and the
    filled-contour drawing in one step.
-   `bins = 12` sets how many contour bands to draw — more bands, finer
    gradation.
-   `coord_sf()` keeps the aspect ratio correct so the map isn't
    stretched.

We can add the neighbourhood boundaries for context. Because the
boundaries are an `sf` object and our density layer isn't, we need
`inherit.aes = FALSE` so `geom_sf` doesn't try to reuse the `x`/`y`
aesthetics:

```{r}
ggplot(listings_xy, aes(x = X, y = Y)) +
  geom_density_2d_filled(bins = 12) +
  geom_sf(data = BA_proj, fill = NA, colour = "white",
          linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_viridis_d(name = "Density") +
  coord_sf() +
  ggtitle("KDE of Airbnbs in Buenos Aires") +
  theme_void()
```

### Changing the bandwidth

The `adjust` argument multiplies the default bandwidth: values below 1
give a spikier surface, above 1 a smoother one. Compare these two
against the default above:

```{r}
# Half the default bandwidth -- more local detail
ggplot(listings_xy, aes(x = X, y = Y)) +
  geom_density_2d_filled(bins = 12, adjust = 0.5) +
  scale_fill_viridis_d(name = "Density") +
  coord_sf() +
  ggtitle("Bandwidth: adjust = 0.5") +
  theme_void()
```

```{r}
# Double the default bandwidth -- much smoother
ggplot(listings_xy, aes(x = X, y = Y)) +
  geom_density_2d_filled(bins = 12, adjust = 2) +
  scale_fill_viridis_d(name = "Density") +
  coord_sf() +
  ggtitle("Bandwidth: adjust = 2") +
  theme_void()
```

::: {.callout-warning}
## These are working maps, not finished ones

Everything we've mapped so far is deliberately rough — the point has
been to see what the method does, not to produce something
publication-ready. Look closely and you'll spot plenty that still needs
fixing:

-   The legend labels are raw contour brackets like `(0.0e+00, 5.0e-09]`
    — technically correct, meaningless to a reader
-   No scale bar, no north arrow, no source credit
-   `"Density"` as a legend title says nothing about *what* is dense, or
    in what units
-   No basemap or boundaries for geographic context on most of them
-   Titles are debugging notes to ourselves (`"Bandwidth: adjust = 2"`),
    not something you'd caption in a report

Before any of these went into a piece of written work, you'd want to
clean all of that up — much like we did in the
[choropleths lab](mapvectorR.html), where we built up a final map with a
proper title, palette, north arrow, scale bar and source.
:::

::: {.callout-tip}
## Have a think

Which of the three bandwidths would you actually put in a report, and
why? There's no right answer — it depends on whether you're trying to
show *where the main concentrations are* or *how fine-grained the
clustering gets*.
:::

::: {.callout-note}
## Going further: the `eks` package

`geom_density_2d_filled()` is quick and needs nothing extra, but it has
limits. It doesn't know that Buenos Aires has a boundary, so it will
happily smear density out over the river; and `adjust` is a hand-tuned
multiplier rather than a statistically-chosen bandwidth.

If you need more control, the [`eks`](https://cran.r-project.org/web/packages/eks/vignettes/tidysf_kde.html)
package works natively with `sf` objects and returns contours as `sf`,
so they slot straight into normal `geom_sf()` mapping:

```{r}
#| eval: false
skde <- eks::st_kde(listings_proj)

ggplot() +
  geom_sf(data = eks::st_get_contour(skde), aes(fill = contlabel), colour = NA) +
  geom_sf(data = BA_proj, fill = NA, colour = "white", linewidth = 0.2) +
  scale_fill_viridis_d(name = "Density") +
  theme_void()
```

The other main option is [`spatstat`](https://spatstat.org/), which is
the most powerful of the three: it handles observation windows properly
(so density is only estimated *inside* your study area) and offers
cross-validated bandwidth selection via `bw.diggle()` and friends. It's
also the most involved — you have to convert your data into its own
`ppp` and `owin` object classes first.

**This is a good moment to practise reading documentation.** We're not
going to walk through either package line by line. If you want to use
them, go and look at their vignettes, work out what the functions
expect, and try it — that skill will serve you far better over the rest
of this course than us handing you the code.
:::

## Cluster of points (DBSCAN)

Partitioning methods (K-means, PAM clustering) and hierarchical
clustering are suitable for finding spherical-shaped clusters or convex
clusters. In other words, they work well for compact and well separated
clusters. Moreover, they are also severely affected by the presence of
noise and outliers in the data.

Unfortunately, real life data can contain: i) clusters of arbitrary
shape ii) many outliers and noise.

In this section, we will learn a method to identify clusters of points,
based on their density across space. To do this, we will use the widely
used `DBSCAN` algorithm. For this method, a cluster is a concentration
of at least `m` points, each of them within a distance of `r` of at
least another point in the cluster. Points in the dataset are then
divided into three categories:

- *Noise*, for those points outside a cluster.
- *Cores*, for those points inside a cluster whith at least `m` points
  in the cluster within distance `r`.
- *Borders* for points inside a cluster with less than `m` other points
  in the cluster within distance `r`.

Both `m` and `r` need to be prespecified by the user before running
`DBSCAN`. This is a critical point, as their value can influence
significantly the final result. Before exploring this in greater depth,
let us get a first run at computing `DBSCAN`, using the `dbscan`
package, which implements the algorithm.

### Data preparation for DBSCAN

`DBSCAN`'s `eps` parameter is a real-world distance, so we need our
points expressed in a projected CRS with metres as units — not raw
longitude/latitude, and not a standardised/scaled version of them, since
neither has a meaningful "distance" interpretation. We already have
exactly that: `listings_proj`, created earlier for the hex-binning
section, in EPSG:22176 (metres). We just need to pull the X/Y
coordinates out into a plain matrix for `dbscan`:

```{r}
coords <- st_coordinates(listings_proj)
head(coords)
```

### Computing DBSCAN using the dbscan package

First, we set the 'random seed', which means that the results will
always be the same when running the following commands, since the random
aspects of the algorithms are controlled.

```{r}
set.seed(123456789)
```

Run the DBSCAN algorithm, specifying:

- `eps`: 'epsilon', radius (in metres, since `coords` is projected) of
  the 'epsilon neighborhood' (the maximum point-to-point distance for
  considering two points to be in the same cluster)
- `minPts`: the minimum number of points required to be in the 'epsilon
  neighborhoods' of core points (including the point itself).

We decide to consider a cluster of airbnb with more than 50 airbnbs
within 100 metres from them, hence we set the two parameters accordingly
— matching the Python lab's first run exactly:

```{r}
db <- dbscan::dbscan(coords, eps = 100, minPts = 50)
db
```

Use the results from applying `dbscan` to plot the example data once
more, coloring points according to which cluster `dbscan` grouped each
point in. In the `dbscan` results, cluster group '0', plotted below in
black, indicates 'noise points'. A 'noise point' is one which isn't
close enough to ('minPts' - 1) number of other points to be considered
part of any cluster.

```{r, warning=FALSE}
factoextra::fviz_cluster(db, coords, stand = FALSE, geom = "point", pointsize = 0.5) + 
  theme_minimal() +
  ggtitle("DBSCAN: eps = 100m, minPts = 50")
```

The algorithm is able to identify a few clusters with a high density of
airbnbs. However, this is all contingent on the parameters we
arbitrarily set. Depending on the maximum radius (`eps`) we set, we will
pick one type of cluster or another: a higher (lower) radius will
translate into less (more) local clusters. Equally, the minimum number
of points required for a cluster (`minPts`) will affect the implicit
size of the cluster.

For an illustration of this, let's run through a case with very
different parameter values: a larger radius (250m) and a smaller minimum
number of points (10) — matching the Python lab's second run:

```{r}
db2 <- dbscan::dbscan(coords, eps = 250, minPts = 10)

factoextra::fviz_cluster(db2, coords, stand = FALSE, geom = "point", pointsize = 0.5) + 
  theme_minimal() +
  ggtitle("DBSCAN: eps = 250m, minPts = 10")
```

The output is now very different, isn't it? This exemplifies how
different parameters can give rise to substantially different outcomes,
even if the same data and algorithm are applied.

The `dbscan` algorithm is very sensitive to changes to the `epsilon` and
`minPts` values. Smaller `epsilons` leads to definition of sparser
clusters as noise while larger epsilon sizes may make denser clusters to
be merged.

### Determining the optimal eps value

Rather than picking `eps` arbitrarily, think of every point and its
distance from its nearest neighbours. We can use a k-nearest-neighbour
distance matrix to compute this, with a specified value of `k`
corresponding to `minPts`.

We then plot these distances in ascending order, with the aim of finding
the "knee" — the point where a sharp change occurs along the curve.
Points below the knee are close enough together to plausibly be part of
a cluster; points above it start looking more like noise. That knee is a
reasonable candidate value for `eps`.

We can use the `kNNdistplot()` function from the `dbscan` package — here
with `k = 50`, matching the `minPts` from our first run:

```{r}
dbscan::kNNdistplot(coords, k = 50)
```

Look for where the curve visibly bends upward — that distance (in
metres, since `coords` is projected) is a defensible choice of `eps` for
this `minPts`.

### A more robust alternative: HDBSCAN

Python's version of this lab uses A-DBSCAN, an ensemble/adaptive variant
of DBSCAN from the `esda` package that isn't available in R. The closest
equivalent available here is **HDBSCAN** (Hierarchical DBSCAN), which
shares A-DBSCAN's core motivation — handling clusters of varying density
more robustly than vanilla DBSCAN — even though the two algorithms work
differently under the hood. `HDBSCAN` only needs `minPts`; it works out
an appropriate `eps` internally for each region of the data rather than
using one fixed radius everywhere:

```{r}
hdb <- dbscan::hdbscan(coords, minPts = 50)
hdb

factoextra::fviz_cluster(list(data = coords, cluster = hdb$cluster), 
                          stand = FALSE, geom = "point", pointsize = 0.5) +
  theme_minimal() +
  ggtitle("HDBSCAN")
```

## Resources

- [DBSCAN example in
  R](https://www.kaggle.com/code/pmcgovern/dbscan-example-in-r)

- [`spatstat` documentation](https://spatstat.org/) — the most powerful
  R option for point pattern analysis, if you want to go beyond this lab

- [`eks` documentation](https://cran.r-project.org/web/packages/eks/vignettes/tidysf_kde.html)
  — `sf`-native kernel density estimation

- [`ggplot2` 2D density docs](https://ggplot2.tidyverse.org/reference/geom_density_2d.html)
  — the `geom_density_2d_filled()` function used in this lab

- [`dbscan` package
  documentation](https://cran.r-project.org/web/packages/dbscan/dbscan.pdf),
  including `hdbscan()`

- [Arribas-Bel et al.,
  2019](https://www.sciencedirect.com/science/article/pii/S0094119019300944)

- [Point Pattern
  Analysis](https://geographicdata.science/book/notebooks/08_point_pattern_analysis.html)
