# Do-It-Yourself {#sec-pointsdiy .unnumbered}

::: {.panel-tabset group="language"}
## R

```{r message=FALSE}
library(sf)
library(dplyr)
library(ggplot2)
```

## Python

```{python}
import pandas as pd
import geopandas as gpd
import contextily as cx
import matplotlib.pyplot as plt
```
:::

## Task I: Photographs in Tokyo

In this task, you will explore patterns in the distribution of the location of photos. We are going to dip our toes in the lake of point data by looking at a sample of geo-referenced photographs in Tokyo. The dataset comes from the [GDS Python Book](https://geographicdata.science/) and contains 10,000 photographs voluntarily uploaded to the Flickr service.

::: callout-note
We read both files below straight from GitHub, so there is nothing to download — but you do need an internet connection when you run these cells.
:::

### Photos

::: {.panel-tabset group="language"}
## R

```{r}
# Read the CSV file into a data frame
tokyo <- read.csv("https://raw.githubusercontent.com/gdsbook/book/master/data/tokyo/tokyo_clean.csv")

head(tokyo)
```

## Python

```{python}
# Read the CSV file into a data frame
tokyo = pd.read_csv("https://raw.githubusercontent.com/gdsbook/book/master/data/tokyo/tokyo_clean.csv")

tokyo.info()
```
:::

The table gives you, for each photo: the `user_id` of the Flickr user who took it, its `longitude`/`latitude` (WGS84), the `date_taken`, a link to the photo page, and `x`/`y` — the same location already projected to Web Mercator.

::: callout-important
`tokyo` is a plain table, not a spatial object yet. Before you can map it you'll need to turn those coordinates into geometries — `st_as_sf()` in R, `gpd.points_from_xy()` in Python — exactly as you did at the start of the lab.
:::

### Administrative areas

::: {.panel-tabset group="language"}
## R

```{r}
url <- "https://raw.githubusercontent.com/darribas/gds_course/master/content/data/tokyo_admin_boundaries.geojson"

tokyo_areas <- sf::st_read(url) # import

head(tokyo_areas)
```

## Python

```{python}
url = "https://raw.githubusercontent.com/darribas/gds_course/master/content/data/tokyo_admin_boundaries.geojson"

tokyo_areas = gpd.read_file(url) # import

# Display the first few rows of the GeoDataFrame
tokyo_areas.head()
```
:::

This gives you 39 administrative areas. The column you'll want for joining and labelling is `NAME_2`.

With these at hand, get to work with the following challenges:

1.  Create a Hex binning map of the photos

2.  Compute and display a kernel density estimate (KDE) of the distribution of the photos

3.  Using the area layer:

    1.  Obtain a count of property by area (no the the area name is present in the property table)

    2.  Create a raw count choropleth

    3.  Create a choropleth of the density of properties by polygon

## Task II: Clusters of Indian cities

For this one, we are going to use a dataset on the location of populated places from [Natural Earth](https://www.naturalearthdata.com/). The original table covers the entire world so, to get it ready for you to work on it, we need to prepare it:

::: {.panel-tabset group="language"}
## R

```{r warning=FALSE}
url <- "https://raw.githubusercontent.com/nvkelso/natural-earth-vector/master/geojson/ne_50m_populated_places_simple.geojson"

data <- sf::st_read(url) # import
data <- st_make_valid(data)

plot(data$geometry) # plot to check all is well
```

## Python

```{python}
url_popplaces = (
    "https://raw.githubusercontent.com/nvkelso/natural-earth-vector/"
    "master/geojson/ne_50m_populated_places_simple.geojson"
)

# Read GeoJSON data directly from the URL
populated_places = gpd.read_file(url_popplaces)

# Plotting the data
fig, ax = plt.subplots(figsize=(12, 8))
populated_places.plot(ax=ax, marker='o', color='black', markersize=10, alpha=0.5)
ax.set_title('Populated Places of the World')
plt.show()
```
:::

Note the code cell above requires internet connectivity.

Now we subset to India only — this leaves us with 71 places to work with:

::: {.panel-tabset group="language"}
## R

```{r}
places <- data %>%
  filter(adm0name == "India")
```

## Python

```{python}
places = populated_places.query("adm0name == 'India'")
```
:::

By default, place locations come expressed in longitude and latitude. Because you will be working with distances, it makes sense to convert the table into a system expressed in metres. For India, this can be the "Kalianpur 1975 / India zone I" (EPSG:24378) projection.

::: {.panel-tabset group="language"}
## R

```{r}
places_m <- st_transform(places, crs = 24378)
```

## Python

```{python}
places_m = places.to_crs(epsg=24378)

```
:::

This is what we have to work with then:

::: {.panel-tabset group="language"}
## R

```{r warning=FALSE}
ggplot() +
  geom_sf(data = places_m, aes(size = pop_max), colour = "goldenrod", alpha = 0.7) +
  scale_size_continuous(name = "Population", range = c(1, 9), labels = scales::comma) +
  ggtitle("Populated places in India") +
  theme_void()
```

## Python

```{python}
ax = places_m.plot(
    color="xkcd:bright yellow", markersize=places_m["pop_max"] / 50000,
    alpha=0.7, figsize=(9, 9)
)
cx.add_basemap(
    ax,
    crs=places_m.crs,
    source=cx.providers.Esri.WorldGrayCanvas
)
ax.set_title("Populated places in India")
ax.set_axis_off()
plt.show()
```
:::

With this at hand, get to work:

1.  Use the DBSCAN algorithm to identify clusters

2.  Start with the following parameters: at least five cities for a cluster (min_samples) and a maximum of 1,000Km (eps)

3.  Obtain the clusters and plot them on a map. Does it pick up any interesting pattern?

4.  Based on the results above, tweak the values of both parameters to find a cluster of southern cities, and another one of cities in the North around New Dehli
